---
title: "Preliminary analysis"
author:
  - "trnguyen@ukaachen.de"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
    outputdir: NA
    inputSeurat: NA
---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.width=20, fig.height=12}
#####----------------------------------------------------------------------#####
#
# trnguyen@ukaachen.de
#
#####----------------------------------------------------------------------#####
source("src/helper_functions.R")
source("src/import_libraries.R")

path.to.save.output <- params$outputdir
path.to.input.s.obj <- params$inputSeurat

dir.create(path.to.save.output, showWarnings = FALSE, recursive = TRUE)
s.obj <- readRDS(path.to.input.s.obj)

# Tableau 20 palette (10 bold + 10 medium pairs, visually distinct)
tableau20 <- c(
  "#4E79A7", "#A0CBE8", "#F28E2B", "#FFBE7D", "#59A14F",
  "#8CD17D", "#B6992D", "#F1CE63", "#499894", "#86BCB6",
  "#E15759", "#FF9D9A", "#79706E", "#BAB0AC", "#D37295",
  "#FABFD2", "#B07AA1", "#D4A6C8", "#9D7660", "#D7B5A6"
)

cluster_colors <- function(n) {
  rep_len(tableau20, n)
}

```

# Preprocessing and QC
## Raw data Quality control  {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (plot.name in names(s.obj@misc$all.QC)){
 
    cat('###',plot.name,'{.unlisted .unnumbered} \n')
    
    ##### 
    # plots or tables that we want to show in tabs
    #####
    print(s.obj@misc$all.QC[plot.name])
    cat(' \n \n')
}
```

## Ambient RNA background correction 

### DecontX clusters {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (plot.name in names(s.obj@misc$ambient.cluster.RNA.plot)){
 
    cat('####',plot.name,'{.unlisted .unnumbered} \n')
    
    ##### 
    # plots or tables that we want to show in tabs
    #####
    print(s.obj@misc$ambient.cluster.RNA.plot[plot.name])
    cat(' \n \n')
}
```

### Contamination level in each sample

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
ggplot(s.obj@meta.data, aes(x=name, y=AmbientRNA)) + 
  geom_boxplot()

```


### Ambient RNA contamination level {.tabset}

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (plot.name in names(s.obj@misc$ambient.contamination.plot)){
 
    cat('####',plot.name,'{.unlisted .unnumbered} \n')
    
    ##### 
    # plots or tables that we want to show in tabs
    #####
    print(s.obj@misc$ambient.contamination.plot[plot.name])
    cat(' \n \n')
}
```


## Descriptive statistics and filtering threshold {.tabset}
This section is devoted to the descriptive statistics of the following varialbes: `nFeature_RNA, nCount_RNA, percent.mt, percent.ribo`. 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (plot.item in c("nFeature_RNA", "nCount_RNA", "percent.mt", "percent.ribo")){
  boxplot <- s.obj@meta.data %>% ggplot(aes_string(x = "name", y = plot.item)) +
    geom_boxplot(outlier.colour="black", outlier.shape=16,
               outlier.size=2, notch=FALSE) +
    ggtitle(sprintf("Boxplot: Distribution of %s in each dataset", plot.item))
  cat('###', plot.item,'{.unlisted .unnumbered} \n')
    
    ##### 
    # plots or tables that we want to show in tabs
    #####
  
  print(boxplot)
  
  cat(' \n \n')
}
```



## Descriptive statistics + UMAP {.tabset}

### % Mitochondrial
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
FeaturePlot(object = s.obj, reduction = "RNA_UMAP", feature = "percent.mt", label = TRUE, label.size = 8, pt.size = 0.5, label.color = "red", )
```

### % Ribosome
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
FeaturePlot(object = s.obj, reduction = "RNA_UMAP", feature = "percent.ribo", label = TRUE, label.size = 8, pt.size = 0.5, label.color = "red", )
```

### % nCount RNA
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
FeaturePlot(object = s.obj, reduction = "RNA_UMAP", feature = "nCount_RNA", label = TRUE, label.size = 8, pt.size = 0.5, label.color = "red", )
```

### % nGenes 
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
FeaturePlot(object = s.obj, reduction = "RNA_UMAP", feature = "nFeature_RNA", label = TRUE, label.size = 8, pt.size = 0.5, label.color = "red", )
```

## Cell cycle scoring {.tabset}

### Cell cycle, split by Phase
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(s.obj, reduction = "RNA_UMAP", split.by = "Phase")
```

### Cell cycle, group by Phase
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
DimPlot(s.obj, reduction = "RNA_UMAP", group.by = "Phase", label = TRUE, label.size = 8, pt.size = 0.5, label.box = TRUE, repel = TRUE)
```

### PCA, cell cycle, group by Phase
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
all.genes <- rownames(x = s.obj)
s.genes <- paste0("^", cc.genes$s.genes, "$", collapse = "|")
s.genes <- all.genes[grepl(s.genes, all.genes, ignore.case = TRUE)]
g2m.genes <- paste0("^", cc.genes$g2m.genes, "$", collapse = "|")
g2m.genes <- all.genes[grepl(g2m.genes, all.genes, ignore.case = TRUE)]
s.obj <- RunPCA(s.obj, features = c(s.genes, g2m.genes), nfeatures.print = 10, reduction.name="CELLCYCLED_PCA")

DimPlot(s.obj, reduction = "CELLCYCLED_PCA", group.by = "Phase", pt.size = 1)
```

# Dimension reduction with UMAP
## UMAP: all clusters in the sample
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
# set cluster identity and compute palette size once
cluster.col <- grep("\\.cluster\\.", colnames(s.obj@meta.data), value = TRUE)[1]
if (!is.na(cluster.col)) Idents(s.obj) <- cluster.col
n.clusters <- length(unique(Idents(s.obj)))
DimPlot(object = s.obj, reduction = "RNA_UMAP", label = TRUE, label.box = TRUE, label.size = 10, repel = TRUE) +
  scale_color_manual(values = cluster_colors(n.clusters)) +
  ggtitle(sprintf("UMAP Sample %s: All clusters", sample.id)) +
  theme(axis.text = element_text(size=20, face = "bold"),
        axis.title = element_text(size=20, face = "bold"),
        title = element_text(size = 20, face = "bold"))
```

## UMAP: ambient RNA contamination level on UMAP
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
FeaturePlot(object = s.obj, reduction = "RNA_UMAP", feature = "AmbientRNA")
```

## Violin plot: Ambient RNA contamination level
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
VlnPlot(object = s.obj, feature = "AmbientRNA")
```

# Cluster marker genes
Identify differentially expressed genes in each cluster. 
```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.height=10, fig.width=18}
Idents(s.obj) <- "RNA_PCA.cluster.0.5"
if (file.exists(file.path(path.to.save.output, "DE_cluster_marker_genes.rds")) == FALSE){
  DefaultAssay(s.obj) <- "SCT"
  s.obj <- PrepSCTFindMarkers(s.obj)
  cluster.markers <- FindAllMarkers(object = s.obj, assay = "SCT", test.use = "wilcox", slot = "data", min.pct = 0.5)
  cluster.markers <- subset(cluster.markers, cluster.markers$p_val_adj < 0.05 & cluster.markers$avg_log2FC > 0)
  saveRDS(cluster.markers, file.path(path.to.save.output, "DE_cluster_marker_genes.rds"))
} else {
  cluster.markers <- readRDS(file.path(path.to.save.output, "DE_cluster_marker_genes.rds"))
}
```

## Feature plot {.tabset}
```{r echo=FALSE, fig.height=12, fig.width=20, message=FALSE, warning=FALSE, results='asis'}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- FeaturePlot(object = s.obj, reduction = "RNA_UMAP", features = head(tmp.cluster.markers, 9)$gene, ncol = 3, label = TRUE, pt.size = 0.5, label.size = 5, label.color = "red", slot = "data")  
  print(p)
  cat("\n \n")
}
```

## Dot plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- DotPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

## Violin plot{.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
for (cluster.id in sort(unique(cluster.markers$cluster))){
  DefaultAssay(s.obj) <- "SCT"
  cat(sprintf("### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- VlnPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene, slot = "data") &
    scale_fill_manual(values = cluster_colors(n.clusters))
  print(p)
  cat("\n \n")
}
```

## Full tables of DE genes {.tabset}
```{r echo=FALSE, results='asis', include=FALSE}
cluster.markers %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
all.cluster.ids <- sort(unique(s.obj$RNA_PCA.cluster.0.5))
for (cluster.id in all.cluster.ids){
  tmp.table <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  cat(paste("\n\n### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.table, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")  
}
```

# Number of cells in each cluster in each sample
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
cell.count.in.clusters <- table(s.obj$name, s.obj$RNA_PCA.cluster.0.5) %>% as.data.frame() %>%
  pivot_wider(names_from = "Var2", values_from = "Freq") 
cell.count.in.clusters %>% create_dt()
```