---
title: "CellChat analysis ŌĆö `r params$sample.id`"
author: "BioFlow Portal"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
  sample.id: "ALL"
  path.to.s.obj: NA
  path.to.save.output: NA
  filter10cells: "NoFilter"
  reduction.name: "umap"
  cluster.name: "seurat_clusters"
  input.spec: "Human"
---

```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'});
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>

```{r setup, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
options(stringsAsFactors = FALSE)

for (pkg in c("CellChat", "Seurat", "dplyr", "stringr", "ggplot2", "patchwork", "DT", "tibble", "hash")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    if (pkg == "CellChat") {
      if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
      BiocManager::install(c("BiocNeighbors", "ComplexHeatmap"), ask = FALSE, update = FALSE)
      devtools::install_github("jinworks/CellChat", upgrade = "never")
    } else {
      install.packages(pkg, repos = "https://cloud.r-project.org")
    }
  }
}

suppressPackageStartupMessages({
  library(CellChat)
  library(Seurat)
  library(dplyr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(DT)
  library(tibble)
  library(hash)
})

# Inline helper
create_dt <- function(x) {
  DT::datatable(x,
    extensions = "Buttons", filter = "top",
    options = list(
      dom = "Blfrtip",
      buttons = c("copy", "csv", "excel"),
      lengthMenu = list(c(10, 25, 50, -1), c(10, 25, 50, "All"))
    ))
}
```

```{r load, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
sample.id           <- params$sample.id
path.to.s.obj       <- params$path.to.s.obj
path.to.save.output <- params$path.to.save.output
filter10cells       <- params$filter10cells
reduction.name      <- params$reduction.name
cluster.name        <- params$cluster.name
input.spec          <- params$input.spec

dir.create(path.to.save.output, showWarnings = FALSE, recursive = TRUE)
s.obj <- readRDS(path.to.s.obj)
s.obj.original <- s.obj

# Subset to a specific sample if requested
if (!is.na(sample.id) && sample.id != "ALL" && "label" %in% colnames(s.obj@meta.data)) {
  s.obj <- subset(s.obj, label == sample.id)
}
```

# Dataset overview

## All cells (UMAP)
```{r umap-all, echo=FALSE, warning=FALSE, message=FALSE, fig.width=14, fig.height=10}
if (reduction.name %in% names(s.obj.original@reductions)) {
  DimPlot(s.obj.original, reduction = reduction.name, label = TRUE, label.box = TRUE, repel = TRUE, group.by = cluster.name)
} else {
  message(sprintf("Reduction '%s' not found. Available: %s", reduction.name, paste(names(s.obj.original@reductions), collapse = ", ")))
}
```

## Selected subset
```{r umap-subset, echo=FALSE, warning=FALSE, message=FALSE, fig.width=14, fig.height=10}
if (reduction.name %in% names(s.obj@reductions)) {
  DimPlot(s.obj, reduction = reduction.name, label = TRUE, label.box = TRUE, repel = TRUE, group.by = cluster.name) +
    ggtitle(sprintf("Sample: %s", sample.id))
}
```

# CellChat setup

```{r cellchat-setup, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
# Detect assay: prefer SCT, fall back to RNA
use.assay <- if ("SCT" %in% names(s.obj@assays)) "SCT" else "RNA"
data.input <- GetAssayData(object = s.obj, slot = "data", assay = use.assay)

meta.data <- s.obj@meta.data %>% rownames_to_column("barcode")
meta.data$cluster <- sapply(meta.data[[cluster.name]], function(x) sprintf("cluster_%s", x))

check.type <- sample(unique(s.obj@meta.data[[cluster.name]]), 1)
if (typeof(check.type) %in% c("double", "integer")) {
  all.idx <- sort(as.numeric(unique(s.obj@meta.data[[cluster.name]])))
  meta.data$cluster <- factor(meta.data$cluster, levels = sapply(all.idx, function(i) sprintf("cluster_%s", i)))
} else {
  meta.data$cluster <- meta.data[[cluster.name]]
}
meta.data <- meta.data %>% column_to_rownames("barcode")

cellchat <- createCellChat(object = data.input, meta = meta.data, group.by = "cluster")
cellchat <- addMeta(cellchat, meta = meta.data)
cellchat <- setIdent(cellchat, ident.use = "cluster")
groupSize <- as.data.frame(table(cellchat@idents))
create_dt(groupSize)
```

```{r db-setup, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
CellChatDB <- if (input.spec == "Human") CellChatDB.human else CellChatDB.mouse
showDatabaseCategory(CellChatDB)
```

```{r run-cellchat, echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.width=14, fig.height=10}
cache.file <- file.path(path.to.save.output,
  sprintf("CellChat_object.%s.%s.rds", str_replace_all(sample.id, " ", "_"), filter10cells))

if (!file.exists(cache.file)) {
  CellChatDB.use <- CellChatDB
  cellchat@DB <- CellChatDB.use
  cellchat <- subsetData(cellchat)
  future::plan("multicore", workers = 4)
  cellchat <- identifyOverExpressedGenes(cellchat)
  cellchat <- identifyOverExpressedInteractions(cellchat)
  cellchat <- computeCommunProb(cellchat)
  if (filter10cells == "Filter10") {
    cellchat <- filterCommunication(cellchat, min.cells = 10)
  }
  cellchat <- computeCommunProbPathway(cellchat)
  cellchat <- aggregateNet(cellchat)
  saveRDS(cellchat, cache.file)
} else {
  cellchat <- readRDS(cache.file)
}
```

# Signal tables {.tabset}

## Counts
Number of interactions between cell groups (row ŌåÆ source, column ŌåÆ target).
```{r mat-count, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
mat.count  <- cellchat@net$count
mat.weight <- cellchat@net$weight
create_dt(as.data.frame(mat.count))
```

## Weights
Total reaction strength (weights proportional to interaction strength).
```{r mat-weight, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
create_dt(as.data.frame(mat.weight))
```

# Aggregated network

## Number of interactions {.tabset}

### Circle
```{r circle-count, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=10, fig.height=10}
groupSize <- as.numeric(table(cellchat@idents))
par(mfrow = c(1, 1), xpd = TRUE)
netVisual_circle(mat.count, vertex.weight = groupSize, weight.scale = TRUE,
                 label.edge = FALSE, title.name = "Number of interactions")
```

### Heatmap
```{r heatmap-count, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=10, fig.height=8}
netVisual_heatmap(cellchat, measure = "count", color.heatmap = "Blues")
```

## Interaction strength {.tabset}

### Circle
```{r circle-weight, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=10, fig.height=10}
par(mfrow = c(1, 1), xpd = TRUE)
netVisual_circle(mat.weight, vertex.weight = groupSize, weight.scale = TRUE,
                 label.edge = FALSE, title.name = "Interaction weights/strength")
```

### Heatmap
```{r heatmap-weight, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=10, fig.height=8}
netVisual_heatmap(cellchat, measure = "weight", color.heatmap = "Reds")
```

# Sent signals per cluster

```{r sent-signals, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(2, 2), xpd = TRUE)
for (i in seq_len(nrow(mat.weight))) {
  mat2 <- matrix(0, nrow = nrow(mat.weight), ncol = ncol(mat.weight), dimnames = dimnames(mat.weight))
  mat2[i, ] <- mat.weight[i, ]
  mat2[i, i] <- 0
  strength <- sort(mat2[i, ], decreasing = TRUE)[1:min(5, ncol(mat2))]
  strength <- strength[strength != 0]
  names(strength) <- str_replace(names(strength), "cluster_", "")
  title <- sprintf("Top signals from %s ŌåÆ %s", rownames(mat.weight)[i], paste(names(strength), collapse = ", "))
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE,
                   edge.weight.max = max(mat.weight), title.name = title, arrow.size = 1)
}
```

# Received signals per cluster

```{r received-signals, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
par(mfrow = c(2, 2), xpd = TRUE)
for (i in seq_len(nrow(mat.weight))) {
  mat2 <- matrix(0, nrow = nrow(mat.weight), ncol = ncol(mat.weight), dimnames = dimnames(mat.weight))
  mat2[, i] <- mat.weight[, i]
  mat2[i, i] <- 0
  strength <- sort(mat2[, i], decreasing = TRUE)[1:min(5, nrow(mat2))]
  strength <- strength[strength != 0]
  names(strength) <- str_replace(names(strength), "cluster_", "")
  title <- sprintf("Top signals to %s from %s", rownames(mat.weight)[i], paste(names(strength), collapse = ", "))
  netVisual_circle(mat2, vertex.weight = groupSize, weight.scale = TRUE,
                   edge.weight.max = max(mat.weight), title.name = title, arrow.size = 1)
}
```

# Pathway-level visualization

```{r pathway-setup, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
all.pathways  <- cellchat@netP$pathways
all.clusters  <- rownames(mat.weight)
pathway.groups <- split(all.pathways, ceiling(seq_along(all.pathways) / 7))
```

Total pathways detected: **`r length(all.pathways)`**

```{r pathway-list, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
cat(paste(all.pathways, collapse = ", "))
```

## Sent signals {.tabset}

```{r sent-pathway, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (cluster.id in all.clusters) {
  cat(sprintf("### %s {.tabset}\n\n", cluster.id))
  for (g in names(pathway.groups)) {
    cat(sprintf("#### Group %s {.tabset}\n\n", g))
    for (pathway in pathway.groups[[g]]) {
      cat(sprintf("##### %s\n\n", pathway))
      tryCatch({
        par(mfrow = c(1, 1), xpd = TRUE)
        netVisual_aggregate(cellchat, signaling = pathway, layout = "circle",
                            sources.use = cluster.id,
                            targets.use = setdiff(all.clusters, cluster.id), arrow.size = 1)
      }, error = function(e) message(sprintf("Skipped %s for %s: %s", pathway, cluster.id, e$message)))
      cat("\n\n")
    }
  }
}
```

## Received signals {.tabset}

```{r received-pathway, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (cluster.id in all.clusters) {
  cat(sprintf("### %s {.tabset}\n\n", cluster.id))
  for (g in names(pathway.groups)) {
    cat(sprintf("#### Group %s {.tabset}\n\n", g))
    for (pathway in pathway.groups[[g]]) {
      cat(sprintf("##### %s\n\n", pathway))
      tryCatch({
        par(mfrow = c(1, 1), xpd = TRUE)
        netVisual_aggregate(cellchat, signaling = pathway, layout = "circle",
                            targets.use = cluster.id,
                            sources.use = setdiff(all.clusters, cluster.id), arrow.size = 1)
      }, error = function(e) message(sprintf("Skipped %s for %s: %s", pathway, cluster.id, e$message)))
      cat("\n\n")
    }
  }
}
```

# Pathway heatmaps {.tabset}

```{r pathway-heatmaps, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
for (g in names(pathway.groups)) {
  cat(sprintf("## Group %s {.tabset}\n\n", g))
  for (pathway in pathway.groups[[g]]) {
    cat(sprintf("### %s\n\n", pathway))
    tryCatch({
      p <- netVisual_heatmap(cellchat, signaling = pathway, color.heatmap = "Reds")
      print(p)
    }, error = function(e) {
      p <- ggplot() + ggtitle(sprintf("No data for pathway: %s", pathway))
      print(p)
    })
    cat("\n\n")
  }
}
```
